---
title: "Replication of Utah MaB study"
execute: 
  echo: false
---
In 2022, Tucker Carlson, a popular cable news host, highlighted a controversial COVID-19 policy in Utah:

"In Utah, COVID patients are triaged by a scoring system [for antibody treatments]...]. [R]ace often counts more than physical health. The State of Utah, for example, gives two points simply for not being white [while] congestive heart failure earns one point."

The Utah Department of Health faced criticism for their race-based triage system, which prioritized non-White patients for monoclonal antibodies.

I obtained the underlying data to determine if UDOH's results replicate.



```{r}
#| label: setup
#| message: false

library(tidyverse)
library(janitor)

```


```{r}
#| label: data-load
#| message: false

df <- read_csv("./data/pone.0261508.s001.csv")





df <- df |> 
  clean_names()


df <- df %>%
  mutate(
    sex = as.factor(sex),
    race = as.factor(race),
    ethnicity = as.factor(ethnicity)
  )


df <- df %>% 
        filter(sex == "Female" | sex == "Male") 

df$race <- ifelse(is.na(df$race), "Unknown", as.character(df$race))
df$ethnicity <- ifelse(is.na(df$ethnicity), "Unknown", as.character(df$ethnicity))



df <- cbind(df,model.matrix(~ sex - 1, data = df))
df <- cbind(df,model.matrix(~ race - 1, data = df))
df <- cbind(df,model.matrix(~ ethnicity - 1, data = df))




df <- df %>%
  clean_names()

df$white_nh <- ifelse(df$race == "White" & df$ethnicity == "Not Hispanic, Latino, or Spanish Origin", 1, 0)

df$coc <- ifelse(df$race_black_or_african_american == 1 | df$race_asian ==1 | df$race_american_indian_or_alaska_native == 1 | df$race_multiple == 1 | df$race_native_hawaiian_or_pacific_islander == 1 | df$ethnicity_hispanic_latino_or_spanish_origin == 1, 1, 0)


df <- df %>%
  clean_names()



```


```{r}
df %>% 
        count(age, race, obesity) %>% 
        group_by(age, race) %>% 
        filter(race != "Multiple") %>% 
        mutate(freq = n / sum(n)) %>% 
        mutate(race = as.factor(race)) %>% 
        filter(obesity == 1) %>% 
        ggplot(aes(x=age,y=freq, color = race)) +
        geom_point(alpha = .2) +
        geom_smooth(method = "loess", formula = "y ~ x", se = FALSE) +
        ggtitle("Obesity rates by race (excluding Multiple) and age")
```

```{r}

# 
# # data %>% 
# #         count(age, coc, x28_d_mortality) %>% 
# #         group_by(age, coc) %>% 
# #         mutate(freq = n / sum(n)) %>% 
# #         mutate(coc = as.factor(coc)) %>% 
# #         filter(x28_d_mortality == 1) %>% 
# #         ggplot(aes(x=age,y=freq, color = coc)) +
# #         geom_point() +
# #         geom_smooth(method = "loess") +
# #         ggtitle("Death is quite rare")
# 
# # create an empty list to store the plots
# plot_list2 <- list()
# 
# data$sex_male<-as.factor(data$sex_male)
# # loop through each variable and create the plot
# for (var in vars_to_plot) {
#   p <- data %>%
#     count(!!sym(var), age, hospitalized_14_days, .drop = FALSE) %>%
#     group_by(!!sym(var), age) %>%
#     mutate(freq = n / sum(n)) %>% 
#     filter(hospitalized_14_days == 1) %>% 
#     ggplot(aes(x=age,y=freq, color = !!sym(var))) +
#         geom_point(size = .1) +
#         geom_smooth(method = "loess")+
#         ggtitle(var)+
#         ylim(0,1)+
#         theme(text = element_text(size = 6), axis.title.x=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank(), axis.title.y=element_blank(), axis.text.y=element_blank(), axis.ticks.y=element_blank(), panel.spacing = unit(0.1, "lines"), legend.position = "none")
#   # add the plot to the list
#   plot_list2[[var]] <- p
# }
# data$sex_male<-as.numeric(data$sex_male)
# # create the facet wrap plot
# 
# ggarrange(plotlist = plot_list2, ncol=6, nrow = 4, common.legend = TRUE, legend="bottom") %>% 
# annotate_figure(., top = textGrob("Hospitalization rates by comorbidity and age"))
```








