---
title: "Replication of Utah MaB study"
execute: 
  echo: false
---
Is it fair to deny someone medical treatment because of their race?

During the COVID-19 pandemic, the Utah Department of Health decided to triage a scarce (but effective) treatment for COVID by giving patients a "risk score".  In a decision many civil rights lawyers flagged as illegal, they allocated additional points to patients based on their race.

This garnered national media attention and many lawyers noted that this was illegal.

The below uses the data that UDOH claimed justified the policy. It shows clearly that the risk of COVID-19 hospitalization increases greatly with age (across all comorbidities).  Although there are differences between non-White and White patients, the differences are small compared to the effect of age.  

```{r}
#| label: setup
#| message: false

library(Hmisc)
library(binom)
library(gridExtra)
library(grid)
library(ggpubr)
library(tidyverse)
library(janitor)
library(tidyr)

```


```{r}
#| label: data-load
#| message: false

df <- read_csv("./data/pone.0261508.s001.csv")





df <- df |> 
  clean_names()


df <- df |>
  mutate(
    sex = as.factor(sex),
    race = as.factor(race),
    ethnicity = as.factor(ethnicity)
  )


df <- df |> 
        filter(sex == "Female" | sex == "Male") 

df$race <- ifelse(is.na(df$race), "Unknown", as.character(df$race))
df$ethnicity <- ifelse(is.na(df$ethnicity), "Unknown", as.character(df$ethnicity))



df <- cbind(df,model.matrix(~ sex - 1, data = df))
df <- cbind(df,model.matrix(~ race - 1, data = df))
df <- cbind(df,model.matrix(~ ethnicity - 1, data = df))




df <- df |>
  clean_names()

df$white_nh <- ifelse(df$race == "White" & df$ethnicity == "Not Hispanic, Latino, or Spanish Origin", 1, 0)

df$coc <- ifelse(df$race_black_or_african_american == 1 | df$race_asian ==1 | df$race_american_indian_or_alaska_native == 1 | df$race_multiple == 1 | df$race_native_hawaiian_or_pacific_islander == 1 | df$ethnicity_hispanic_latino_or_spanish_origin == 1, 1, 0)


df <- df |>
  clean_names()



```

```{r}
#| warning: false
df <- df |>
  mutate(across(-c(random, charlson, x28_d_mortality, hospitalized_14_days, age), as.factor))


comorb <- c(
  "sex_male",
  "fever_flg",             
  "cough_flg",             
  "sob_flg",               
  "body_aches_flg",       
  "runny_nose_flg",        
  "decr_smell_flg",        
  "sore_throat_flg",       
  "diarrhea_flg",         
  "diabetes",              
  "hypertension",          
  # "myocardial_infarction", 
  "cardiac_arrhythmia",   
  "chronic_pulmonary",     
  "renal_disease",         
  "chf",                   
  "liver_disease",        
  # "metastatic_carcinoma",  
  "obesity",               
#  "transplant_flag",      
  # "immunosupp",            
  "cerebrovascular",       
  "neuro_disorders",
  "no_comorb"
)


# Define a named vector for the facet labels
facet_labels <- c(
  sex_male = "Sex (Male)",
  fever_flg = "Fever",
  cough_flg = "Cough",
  sob_flg = "Shortness of Breath",
  body_aches_flg = "Body Aches",
  runny_nose_flg = "Runny Nose",
  decr_smell_flg = "Decreased Smell",
  sore_throat_flg = "Sore Throat",
  diarrhea_flg = "Diarrhea",
  diabetes = "Diabetes",
  hypertension = "Hypertension",
  cardiac_arrhythmia = "Cardiac Arrhythmia",
  chronic_pulmonary = "Chronic Pulmonary",
  renal_disease = "Renal Disease",
  chf = "CHF",
  liver_disease = "Liver Disease",
  obesity = "Obesity",
  immunosupp = "Immunosuppression",
  cerebrovascular = "Cerebrovascular",
  neuro_disorders = "Neuro Disorders",
  no_comorb = "No Comorbidity"
)

# Update the ggplot code
df %>%
  mutate(no_comorb = if_all(
    c(
      "diabetes",
      "hypertension",
      "cardiac_arrhythmia",
      "chronic_pulmonary",
      "renal_disease",
      "chf",
      "liver_disease",
      "obesity",
      "immunosupp",
      "cerebrovascular",
      "neuro_disorders"
    ),
    ~ . == 0,
    .names = "no_comorb"
  )) %>%
  mutate(no_comorb = as.integer(no_comorb)) %>%
  mutate(no_comorb = as_factor(no_comorb)) %>%
  pivot_longer(
    cols = any_of(comorb),
    names_to = "comorbidity",
    values_to = "value"
  ) %>%
  filter(value == 1) %>%
  select(hospitalized_14_days, age, comorbidity, coc) %>%
  drop_na() %>%
  ggplot(aes(x = age, y = hospitalized_14_days, color = coc)) +
  geom_point(alpha = 0.01) +
  geom_smooth(method = "loess", formula = y ~ x, se = TRUE) +
  facet_wrap(~ comorbidity, labeller = as_labeller(facet_labels)) +
  theme_minimal() +
  labs(
    title = "Hospitalization rates by comorbidity, age, and race",
    x = "Age",
    y = "Hospitalized (14 Days)",
    color = "Race"
  ) +
  # scale_color_viridis_d(
  #   option = "rocket",
  #   direction = -1,
  #   labels = c("White non-Hispanic", "Not White")
  # ) +
  scale_y_continuous(
    limits = c(0, 1),
    breaks = c(0, 1),
    labels = scales::percent
  ) +
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal"
  )




```
